---
title: "Gene Set Analysis"
output: html_document
date: "2024-07-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library(edgeR)
library(limma)
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
library(ggplot2)
library(ggpubr)

```
## Overview/Notes
Plots used for analysis: 
• Dot plot:number of genes belonging to each GO-term and associated adjusted p-values  

• Enrichment map: Shows overlap between genes in different GO terms

• Category Net plot:Linkages of genes and biological concepts as a network

• Ridge Plot: Density pltos of expression for different gene sets to determine up/downregulated pathways 

```{r, include=FALSE}
#Loading data and creating design matrices for different gene sets
countdata <- read.delim("/Users/dalvinikita/Documents/Drug Response Analysis/countsmatrix.tsv", check.names = FALSE, stringsAsFactors = FALSE)
sampleinfo <- read.delim("/Users/dalvinikita/Documents/Drug Response Analysis/SampleInfo.txt", check.names=FALSE, stringsAsFactors = TRUE)
counts <- DGEList(counts=countdata[,3:14], genes = countdata[,2])
batch <- factor(c(1,1,1,1,1,1,2,2,2,2,2,2))

#TOP1 inhibitors
#calculating avg AUC for TOP1 inhibitors
y<- counts
top1_auc <- cbind(sampleinfo$Irinotecan,sampleinfo$Topotecan)
ave_auc <- rowMeans(top1_auc)
design <- model.matrix(~batch+ave_auc)
#filtering to remove low counts and fitting GLMs 
keep <- filterByExpr(y, design = design)
y <- y[keep, , keep.lib.sizes=FALSE]
y <- calcNormFactors(y)
y <- estimateDisp(y, design, robust=TRUE)
fit <- glmFit(y, design)
lrt<- glmLRT(fit)
sig_genes <- topTags(lrt, p.value=0.05, n=500)
write.table(sig_genes, quote=F, row.names = F, col.names = T, file='/Users/dalvinikita/Documents/Drug Response Analysis/top1_drugs_genes.txt', sep='\t')

#Platinum based drugs
y<- counts
#calculating avg AUC for platinum based drugs
platinum_auc <- cbind(sampleinfo$Carboplatin,sampleinfo$Cisplatin)
platinum_auc <- cbind(platinum_auc,sampleinfo$Oxaliplatin)
ave_auc <- rowMeans(platinum_auc)
design <- model.matrix(~batch+ave_auc)
#filtering to remove low counts and fitting GLMs 
keep <- filterByExpr(y, design = design)
y <- y[keep, , keep.lib.sizes=FALSE]
y <- calcNormFactors(y)
y <- estimateDisp(y, design, robust=TRUE)
fit <- glmFit(y, design)
lrt<- glmLRT(fit)
sig_genes <- topTags(lrt, p.value=0.05, n=500)
write.table(sig_genes, quote=F, row.names = F, col.names = T, file='/Users/dalvinikita/Documents/Drug Response Analysis/plat_drugs_genes.txt', sep='\t')

#Cytotoxic Antibiotics
y<- counts
#calculating avg AUC for cytotoxic antibiotics
cytotoxic_antibiotics <- cbind(sampleinfo$Mitomycin,sampleinfo$Doxorubicin)
ave_auc <- rowMeans(cytotoxic_antibiotics)
design <- model.matrix(~batch+ave_auc)
#filtering to remove low counts and fitting GLMs 
keep <- filterByExpr(y, design = design)
y <- y[keep, , keep.lib.sizes=FALSE]
y <- calcNormFactors(y)
y <- estimateDisp(y, design, robust=TRUE)
fit <- glmFit(y, design)
lrt<- glmLRT(fit, coef=3)
sig_genes <- topTags(lrt, p.value=0.05, n=500)
write.table(sig_genes, quote=F, row.names = F, col.names = T, file='/Users/dalvinikita/Documents/Drug Response Analysis/cytotoxic_antibiotic_genes.txt', sep='\t')

#Taxanes
y<- counts
#calculating avg AUC for taxanes
taxanes <- cbind(sampleinfo$Paclitaxel,sampleinfo$Docetaxel)
ave_auc <- rowMeans(taxanes)
design <- model.matrix(~batch+ave_auc)
#filtering to remove low counts and fitting GLMs 
keep <- filterByExpr(y, design = design)
y <- y[keep, , keep.lib.sizes=FALSE]
y <- calcNormFactors(y)
y <- estimateDisp(y, design, robust=TRUE)
fit <- glmFit(y, design)
lrt<- glmLRT(fit, coef=3)
sig_genes <- topTags(lrt, p.value=0.05, n=500)
write.table(sig_genes, quote=F, row.names = F, col.names = T, file='/Users/dalvinikita/Documents/Drug Response Analysis/taxane_genes.txt', sep='\t')
```
## Analysis of significantly correlated genes across TOP1 inhibitors 
### Significant genes 
```{r}
#loading in data
top1_sig_genes <- read.delim('/Users/dalvinikita/Documents/Drug Response Analysis/top1_drugs_genes.txt', header=T)
head(top1_sig_genes)
```
### Preparing input 
```{r message=FALSE, warning=FALSE}
gene_list <- top1_sig_genes$logFC
names(gene_list)<- top1_sig_genes$genes
gene_list=sort(gene_list, decreasing = TRUE)

top1_gse <- gseGO(geneList=gene_list, 
             ont ="ALL", 
             keyType = "ALIAS", 
             minGSSize = 3, 
             maxGSSize = 800, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = org.Hs.eg.db, 
             pAdjustMethod = "none")
```
### DotPlot 
```{r top1_dotplot, message=FALSE, warning=FALSE}
require(DOSE)
dotplot(top1_gse, showCategory=8,split=".sign",label_format=50,color='p.adjust')+
  facet_grid(.~.sign) +
  theme_dose(font.size = 5)+
  ggtitle("dotplot for TOP1 inhibitors")
```
### Enrichment Map
```{r top1_enrichment_map}
x<- pairwise_termsim(top1_gse)
emapplot(x,layout.params=list(layout='circle'))+
  ggtitle("enrichment map for TOP1 inhibitors")
```
### Category Netplot
```{r top1_cnetPlot}
by_pval <- cnetplot(top1_gse, categorySize="pvalue", color.params=list(foldChange=gene_list,category='deeppink1'))+
  ggtitle("By p-value")

by_all <- cnetplot(top1_gse, node_label="all", color.params=list(category='deeppink1', gene='darkmagenta'), showCategory=10)+
  ggtitle("By both gene category and name")

by_category <- cnetplot(top1_gse, node_label="category", showCategory=10, color.params=list(category='deeppink1', gene='darkmagenta'))+
  ggtitle("By gene category")

by_gene <- cnetplot(top1_gse, node_label="gene", color.params=list(category='deeppink1', gene='darkmagenta'), showCategory=10)+
  ggtitle("By gene name")

ggarrange(by_pval, by_all, by_category, by_gene,ncol=2, nrow=2)
```
### Ridgeplot 
```{r top1__ridgeplot}
ridgeplot(top1_gse, showCategory = 15, label_format =50 )+theme_dose(font.size = 5)+
  ggtitle("ridgeplot for TOP1 inhibitors")
```
## Analysis of significantly correlated genes across platinum compounds
```{r, include=FALSE}
#loading in count data
countdata <- read.delim("/Users/dalvinikita/Documents/Drug Response Analysis/countsmatrix.tsv", check.names = FALSE, stringsAsFactors = FALSE)
sampleinfo <- read.delim("/Users/dalvinikita/Documents/Drug Response Analysis/SampleInfo.txt", check.names=FALSE, stringsAsFactors = TRUE)
y <- DGEList(counts=countdata[,3:14], genes = countdata[,2])

batch <- factor(c(1,1,1,1,1,1,2,2,2,2,2,2))

#calculating avg AUC for platinum based drugs
platinum_auc <- cbind(sampleinfo$Carboplatin,sampleinfo$Cisplatin)
platinum_auc <- cbind(platinum_auc,sampleinfo$Oxaliplatin)
ave_auc <- rowMeans(platinum_auc)
design <- model.matrix(~batch+ave_auc)

#filtering to remove low counts and fitting GLMs 
keep <- filterByExpr(y, design = design)
y <- y[keep, , keep.lib.sizes=FALSE]
y <- calcNormFactors(y)
y <- estimateDisp(y, design, robust=TRUE)
fit <- glmFit(y, design)
lrt<- glmLRT(fit)
sig_genes <- topTags(lrt, p.value=0.05, n=500)
write.table(sig_genes, quote=F, row.names = F, col.names = T, file='/Users/dalvinikita/Documents/Drug Response Analysis/plat_drugs_genes.txt', sep='\t')
```
### Significant genes 
```{r}
#loading in data
plat_sig_genes <- read.delim('/Users/dalvinikita/Documents/Drug Response Analysis/plat_drugs_genes.txt', header=T)
head(plat_sig_genes)
```
### Preparing input 
```{r message=FALSE, warning=FALSE}
gene_list <- plat_sig_genes$logFC
names(gene_list)<- plat_sig_genes$genes
gene_list=sort(gene_list, decreasing = TRUE)

gse <- gseGO(geneList=gene_list, 
             ont ="ALL", 
             keyType = "ALIAS", 
             minGSSize = 3, 
             maxGSSize = 800, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = org.Hs.eg.db, 
             pAdjustMethod = "none")
```
### DotPlot 
```{r platinum_compounds_dotplot, message=FALSE, warning=FALSE}
require(DOSE)
dotplot(gse, showCategory=8, split=".sign",label_format=50,color='p.adjust')+
  facet_grid(.~.sign) +
  theme_dose(font.size = 5)+
  ggtitle("dotplot for platinum based compounds")
```
### Enrichment Map
```{r platinum_compounds_enrichment_map}
x<- pairwise_termsim(gse)
emapplot(x, showCategory=40)+
  ggtitle("enrichment map for platinum based compounds")
```
### Category Netplot
```{r platinum_compounds_cnetPlot}
by_pval <- cnetplot(gse, categorySize="pvalue", color.params=list(foldChange=gene_list,category='deeppink1'))+
  ggtitle("By p-value")

by_all <- cnetplot(gse, node_label="all", color.params=list(category='deeppink1', gene='darkmagenta'), showCategory=10)+
  ggtitle("By both gene category and name")

by_category <- cnetplot(gse, node_label="category", showCategory=10, color.params=list(category='deeppink1', gene='darkmagenta'))+
  ggtitle("By gene category")

by_gene <- cnetplot(gse, node_label="gene", color.params=list(category='deeppink1', gene='darkmagenta'), showCategory=10)+
  ggtitle("By gene name")

ggarrange(by_pval, by_all, by_category, by_gene,ncol=2, nrow=2)
```
### Ridgeplot 
```{r platinum_compounds_ridgeplot}
ridgeplot(gse, showCategory = 15, label_format =50 )+theme_dose(font.size = 5)+
  ggtitle("ridgeplot for platinum based compounds")
```

## Analysis of significantly correlated genes across cytotoxic antibiotics

```{r, include=FALSE}
#loading in count data
countdata <- read.delim("/Users/dalvinikita/Documents/Drug Response Analysis/countsmatrix.tsv", check.names = FALSE, stringsAsFactors = FALSE)
sampleinfo <- read.delim("/Users/dalvinikita/Documents/Drug Response Analysis/SampleInfo.txt", check.names=FALSE, stringsAsFactors = TRUE)
y <- DGEList(counts=countdata[,3:14], genes = countdata[,2])

batch <- factor(c(1,1,1,1,1,1,2,2,2,2,2,2))

#calculating avg AUC for cytotoxic antibiotics
cytotoxic_antibiotics <- cbind(sampleinfo$Mitomycin,sampleinfo$Doxorubicin)
ave_auc <- rowMeans(cytotoxic_antibiotics)
design <- model.matrix(~batch+ave_auc)

#filtering to remove low counts and fitting GLMs 
keep <- filterByExpr(y, design = design)
y <- y[keep, , keep.lib.sizes=FALSE]
y <- calcNormFactors(y)
y <- estimateDisp(y, design, robust=TRUE)
fit <- glmFit(y, design)
lrt<- glmLRT(fit, coef=3)
sig_genes <- topTags(lrt, p.value=0.05, n=500)
write.table(sig_genes, quote=F, row.names = F, col.names = T, file='/Users/dalvinikita/Documents/Drug Response Analysis/cytotoxic_antibiotic_genes.txt', sep='\t')
```
### Significant genes 
```{r}
#loading in data
cytotoxic_antibiotic_genes <- read.delim('/Users/dalvinikita/Documents/Drug Response Analysis/cytotoxic_antibiotic_genes.txt', header=T)
head(cytotoxic_antibiotic_genes)
```
### Preparing input 
```{r message=FALSE, warning=FALSE}
ca_gene_list <- cytotoxic_antibiotic_genes$logFC
names(ca_gene_list)<- cytotoxic_antibiotic_genes$genes
ca_gene_list=sort(ca_gene_list, decreasing = TRUE)

ca_gse <- gseGO(geneList=ca_gene_list, 
             ont ="ALL", 
             keyType = "ALIAS", 
             minGSSize = 3, 
             maxGSSize = 800, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = org.Hs.eg.db, 
             pAdjustMethod = "none")
```
### DotPlot 
```{r cytotoxic_antibiotics_dotplot}
require(DOSE)
dotplot(ca_gse, showCategory=8, split=".sign",label_format=50,color='p.adjust')+
  facet_grid(.~.sign) +
  theme_dose(font.size = 5)+
  ggtitle("dotplot for cytotoxic antibiotics")
```

### Enrichment Map
```{r cytotoxic_antibiotics_enrichment_map}
x<- pairwise_termsim(ca_gse)
emapplot(x, showCategory=40)+
  ggtitle("enrichment map for cytotoxic antibiotics")
```
### Category Netplot
```{r cytotoxic_antibiotics_cnetPlot, message=FALSE, warning=FALSE}
by_pval <- cnetplot(ca_gse, categorySize="pvalue", color.params=list(foldChange=gene_list,category='deeppink1'))+
  ggtitle("By p-value")

by_all <- cnetplot(ca_gse, node_label="all", color.params=list(category='deeppink1', gene='darkmagenta'), showCategory=10)+
  ggtitle("By both gene category and name")

by_category <- cnetplot(ca_gse, node_label="category", showCategory=10, color.params=list(category='deeppink1', gene='darkmagenta'))+
  ggtitle("By gene category")

by_gene <- cnetplot(ca_gse, node_label="gene", color.params=list(category='deeppink1', gene='darkmagenta'), showCategory=10)+
  ggtitle("By gene name")

ggarrange(by_pval, by_all, by_category, by_gene,ncol=2, nrow=2)
```
### Ridgeplot 
```{r cytotoxic_antibiotics_ridgeplot, message=FALSE, warning=FALSE}
ridgeplot(ca_gse, showCategory = 15, label_format =50 )+theme_dose(font.size = 5)+
  ggtitle("ridgeplot for cytotoxic antibiotics")
```
## Analysis of significantly correlated genes across taxanes
```{r, include=FALSE}
#loading in count data
countdata <- read.delim("/Users/dalvinikita/Documents/Drug Response Analysis/countsmatrix.tsv", check.names = FALSE, stringsAsFactors = FALSE)
sampleinfo <- read.delim("/Users/dalvinikita/Documents/Drug Response Analysis/SampleInfo.txt", check.names=FALSE, stringsAsFactors = TRUE)
y <- DGEList(counts=countdata[,3:14], genes = countdata[,2])

batch <- factor(c(1,1,1,1,1,1,2,2,2,2,2,2))

#calculating avg AUC for taxanes
taxanes <- cbind(sampleinfo$Paclitaxel,sampleinfo$Docetaxel)
ave_auc <- rowMeans(taxanes)
design <- model.matrix(~batch+ave_auc)

#filtering to remove low counts and fitting GLMs 
keep <- filterByExpr(y, design = design)
y <- y[keep, , keep.lib.sizes=FALSE]
y <- calcNormFactors(y)
y <- estimateDisp(y, design, robust=TRUE)
fit <- glmFit(y, design)
lrt<- glmLRT(fit, coef=3)
sig_genes <- topTags(lrt, p.value=0.05, n=500)
write.table(sig_genes, quote=F, row.names = F, col.names = T, file='/Users/dalvinikita/Documents/Drug Response Analysis/taxane_genes.txt', sep='\t')
```
### Significant genes 
```{r}
#loading in data
taxane_genes <- read.delim('/Users/dalvinikita/Documents/Drug Response Analysis/taxane_genes.txt', header=T)
head(taxane_genes)
```
### Preparing input 
```{r message=FALSE, warning=FALSE}
tax_gene_list <- taxane_genes$logFC
names(tax_gene_list)<- taxane_genes$genes
tax_gene_list=sort(tax_gene_list, decreasing = TRUE)

tax_gse <- gseGO(geneList=tax_gene_list, 
             ont ="ALL", 
             keyType = "ALIAS", 
             minGSSize = 3, 
             maxGSSize = 800, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = org.Hs.eg.db, 
             pAdjustMethod = "none")
```

### DotPlot 
```{r taxanes_dotplot}
require(DOSE)
dotplot(tax_gse, showCategory=8, split=".sign",label_format=50,color='p.adjust')+
  facet_grid(.~.sign) +
  theme_dose(font.size = 5)+
  ggtitle("dotplot for taxanes")
```
### Enrichment Map
```{r taxanes_enrichment_map}
x<- pairwise_termsim(tax_gse)
emapplot(x, showCategory=40)+
  ggtitle("enrichment map for taxanes")
```
### Category Netplot
```{r taxanes_cnetPlot, warning=FALSE}
by_pval <- cnetplot(tax_gse, categorySize="pvalue", color.params=list(foldChange=gene_list,category='deeppink1'))+
  ggtitle("By p-value")

by_all <- cnetplot(tax_gse, node_label="all", color.params=list(category='deeppink1', gene='darkmagenta'), showCategory=10)+
  ggtitle("By both gene category and name")

by_category <- cnetplot(tax_gse, node_label="category", showCategory=10, color.params=list(category='deeppink1', gene='darkmagenta'))+
  ggtitle("By gene category")

by_gene <- cnetplot(tax_gse, node_label="gene", color.params=list(category='deeppink1', gene='darkmagenta'), showCategory=10)+
  ggtitle("By gene name")

ggarrange(by_pval, by_all, by_category, by_gene,ncol=2, nrow=2)
```
### Ridgeplot 
```{r taxanes_ridgeplot, warning=FALSE}
ridgeplot(tax_gse, showCategory = 15, label_format =50 )+theme_dose(font.size = 5)+
  ggtitle("ridgeplot for taxanes")
```
