---
title: "MDS plots"
output: html_document
date: "2024-06-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(edgeR)
library(limma)
#loading in count data
countdata <- read.delim("/Users/dalvinikita/Documents/Drug Response Analysis/countsmatrix.tsv", check.names = FALSE, stringsAsFactors = FALSE)
```
Checking the count data:
```{r}
head(countdata)
```
Loading in the sample metadata:
```{r}
sampleinfo <- read.delim("/Users/dalvinikita/Documents/Drug Response Analysis/SampleInfo.txt", check.names=FALSE, stringsAsFactors = TRUE)

#assigning "Batch information as a character/factor instead of integer
sampleinfo$Batch<- as.character(sampleinfo$Batch)
sampleinfo$Batch<- factor(c(sampleinfo$Batch))
```
Checking the sample metadata:
```{r}
head(sampleinfo)
```
Putting data it a DGElist object:
```{r}
y <- DGEList(counts=countdata[,3:14], genes = countdata[,2])
head(y)
```

# MDS Plots
Setting up columns to colour code MDS plots based on patient, growth pattern, batch, and media:
```{r}
#assigning colours for MDS plots based on patient, growth patters, batch, and media
col.patient <-c("red", "orange", "yellow","lightgreen", "lightblue", "purple", "magenta")[sampleinfo$PATIENT]
col.pattern <-c("red", "orange", "green")[sampleinfo$PATTERN]
col.batch <-c("red", "orange", "green")[sampleinfo$Batch]
col.media <- c("red", "orange")[sampleinfo$Media]
```
Generating plots:
```{r, echo=FALSE}
#creating MDS plots 
MDS_patient<- plotMDS(y, col=col.patient)
legend("topleft", fill= c("red", "orange", "yellow","lightgreen", "lightblue", "purple", "magenta"), legend=levels(sampleinfo$PATIENT))
title("Patient")

MDS_pattern <- plotMDS(y, col=col.pattern)
legend("topleft", fill= c("red", "orange", "green"), legend=levels(sampleinfo$PATTERN))
title("Growth Pattern")

MDS_batch<- plotMDS(y, col=col.batch)
legend("topleft", fill= c("red", "orange", "green"), legend=levels(sampleinfo$Batch))
title("Batch")

MDS_media<- plotMDS(y, col=col.media)
legend("topleft", fill= c("red", "orange"), legend=levels(sampleinfo$Media))
title("Media")
```

Notes: 

• Samples cluster together based on patient from which the sample was taken from, with ORG71 being the most dissimilar

• Samples do not seem to cluster based on growth pattern or media

• Samples show some clustering based on batch - may need to remove ORG71 from this to get a better idea 


## MDS plot of batch and growth pattern without ORG71
```{r, echo=FALSE}
#remove ORG71
y2<- y[,-c(10)]
sampleinfo2 <- sampleinfo[-c(10),]
col.batch2 <-c("red","orange", "green")[sampleinfo$Batch]
col.pattern2 <-c("red","orange", "green")[sampleinfo$PATTERN]


MDS_pattern2<- plotMDS(y2, col=col.pattern2)
legend("topleft", fill= c("red","orange","green"), legend=levels(sampleinfo2$PATTERN))
title("Growth Pattern")

MDS_batch2<- plotMDS(y2, col=col.batch2)
legend("topleft", fill= c("red","orange","green"), legend=levels(sampleinfo2$Batch))
title("Batch")
```

Notes: 

• Tight clustering of samples that were run together in batch 1 (ORG38, ORG41)

• No clustering based on growth pattern

• Factors to include in design matrix: Patient and Batch




