---
title: "Identifying genes related to response using AUC of pred-probs from ML methods"
output: html_document
date: "2025-06-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
	fig.width = 5,
	fig.height=5
)
```

```{r}
library(pheatmap)
library(ggplot2)
library(edgeR)
library(limma)
library(ggpubr)
library(RColorBrewer)
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
library(DT)
library(EnhancedVolcano)
library(readxl)
```

```{r}
### Functions 

### buttons for datatables 
create_dt <- function(x){
  DT::datatable(x,
                extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                               buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
                               lengthMenu = list(c(10,25,50,-1),
                                                 c(10,25,50,"All"))))
}
```

## Notes

This analysis aims to identify genes related to response in MOC organoids, by using AUC (of predicted probabilities generated from Image Analysis ML models) as a measure of sensitivity/response. 

data location:
• RNAseq data :'/home/ndalvi/RNAseq_STAR/Results/HTseq/collated_counts.tsv'

• AUC data: 'data/AUC_metadata.xlsx'


## Data
Taking a quick look at both the RNA-seq data and AUC data 

RNA-seq data:
```{r}
rawdata <- read.delim2('/Volumes/bioinf/home/ndalvi/RNAseq_STAR/Results/HTseq/collated_counts.tsv', header = TRUE, check.names = FALSE, stringsAsFactors = FALSE)
subset <- c('ENSEMBL_GeneID','ORG-38-ORG','ORG41','ORG-46-ORG','ORG49','ORG55','ORG60','ORG61','ORG64_new','ORG65','ORG66','ORG70','ORG73')
head(rawdata[,subset] )
```

Relevant metadata:
```{r}
metadata <- read_excel("/Users/dalvinikita/Documents/7_GitHub/drug-AUC-analysis/data/relAUC_metadata.xls")
metadata <- metadata[c(-13),]
create_dt(metadata)
```
*Note: We have removed ORG76 from the AUC metadata as there is no corresponding RNAseq, but AUC rankings are based on AUC values that included ORG76. 

## QC and Normalisation

### Design 
Taking a look at the design matrix which includes growth pattern and batch of sequencing:

```{r}
### creating a DGElist object to include organoids with AUC data
organoids <- c('ORG-38-ORG','ORG41','ORG-46-ORG','ORG49','ORG55','ORG60','ORG61','ORG64_new','ORG65','ORG66','ORG70','ORG73')
y<- DGEList(counts=rawdata[,organoids], genes=rawdata[,1])

### tidying up sample names
colnames(y) <- c('ORG38-O','ORG41','ORG46-O','ORG49','ORG55','ORG60','ORG61','ORG64','ORG65','ORG66','ORG70','ORG73')

### tidying up metadata sample names to match counts matrix]
metadata$PATIENT <- c('ORG38-O','ORG41','ORG46-O','ORG49','ORG55','ORG60','ORG61','ORG64','ORG65','ORG66','ORG70','ORG73')
rownames(metadata) <- metadata$PATIENT

### creating factors to add into design matrix
### including pattern and batch
pattern <- as.factor(metadata$Pattern)
batch <- as.factor(metadata$Batch)

#create design matrix to include batch, growth pattern
design <- model.matrix(~0+pattern+batch)
rownames(design) <- metadata$PATIENT
create_dt(design)
```

### Normalisation
Filtering to remove genes with low counts:
```{r, fig.width=7}
#filtering to remove low counts
keep <- filterByExpr(y, design)
print(table(keep))
y <- y[keep, , keep.lib.sizes=TRUE]

#Visualiszing library sizes pre and post normalisation
col <- brewer.pal(32, 'Paired')

par(mfrow=c(1,2))
lcpm <- cpm(y, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="A. Example: Unnormalised data",ylab="Log-cpm")

y <- calcNormFactors(y) 
lcpm <- cpm(y, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="B. Example: Normalised data",ylab="Log-cpm")
```

### MDS plots {.tabset}
```{r, echo=FALSE}
col.pattern <- c('darkorchid3', 'deeppink')[as.factor(metadata$Pattern)]
col.batch <- c('darkorchid3', 'deeppink', 'orange')[as.factor(metadata$Batch)]
```

#### Label
```{r, fig.height=7, fig.width=14}
par(mfrow=c(1,2))
plotMDS(y, col=col.pattern)
title(main="A. Growth Pattern")
legend("topleft",
       fill=c('darkorchid3', 'deeppink'), 
       legend=levels(as.factor(metadata$Pattern)))

plotMDS(y, col=col.batch)
title(main="B. Batch")
legend("topleft",
       fill=c('darkorchid3', 'deeppink','orange'), 
       legend=levels(as.factor(metadata$Batch)))

```

#### Point
```{r, fig.height=7, fig.width=14}
par(mfrow=c(1,2))
plotMDS(y, col=col.pattern, pch=19)
title(main="A. Growth Pattern")
legend("topleft",
       fill=c('darkorchid3', 'deeppink'),
       legend=levels(as.factor(metadata$Pattern)))

plotMDS(y, col=col.batch, pch=19)
title(main="B. Batch")
legend("topleft",
       fill=c('darkorchid3', 'deeppink','orange'),
       legend=levels(as.factor(metadata$Batch)))

```

## Method 1: Fitting a GLM {.tabset}
In this approach, we run 11 separate analyses (for all drugs except Staurosporine) to identify genes associated with changes in AUC values for the respective drug, across our 12 organoid lines. Design matrices for all analysis will include: 

• Growth pattern (INF/EXP)

• Batch of sequencing (1,2,3)

• AUC values for respective drug as a numeric/ continuous variable. 

### 5FU

#### Design Matrix 

Design matrix used for analysis:

```{r}
drug <- "5FU"
counts <- y
auc <- as.numeric(metadata$`5-Fluouracil`)
auc_design <- cbind(design,auc)
create_dt(auc_design)
```

#### Fitting GLM

Top genes related to 5FU AUC: 

```{r}
counts <- counts %>%
  estimateDisp() %>%
  estimateCommonDisp() %>%
  estimateTagwiseDisp()

fit <- glmQLFit(counts, design = auc_design, robust = TRUE)
auc_fit <- glmQLFTest(fit, coef=c('auc'))
genes <- topTags(auc_fit, n=Inf)
create_dt(as.data.frame(genes))

```
*Notes: No genes were found to have FDR < 0.05.*

#### Top genes

Plotting top genes: 

```{r}
top_genes <- c(genes$table$genes)
top_genes <- top_genes[1:5]
auc_values <- auc

for (gene in top_genes) {
  gene <- as.character(gene)
  y$cpm <- cpm(y, log = T) 
  expression_values <- y$cpm[y$genes$genes==gene,]
  plot<- ggplot(data = NULL, mapping = aes(x=auc, y=expression_values, colour = batch)) + 
    geom_point() +
    theme_linedraw() +
    xlab('AUC values') + 
    ylab('log(cpm)') +
    ggtitle(label=gene)
  print(plot)
}
```


### Carboplatin

#### Design Matrix 

Design matrix used for analysis:

```{r}
counts <- y
auc <- as.numeric(metadata$Carboplatin)
auc_design <- cbind(design,auc)
create_dt(auc_design)
```

#### Fitting GLM

Top genes related to Carboplatin AUC: 

```{r}
counts <- counts %>%
  estimateDisp() %>%
  estimateCommonDisp() %>%
  estimateTagwiseDisp()

fit <- glmQLFit(counts, design = auc_design, robust = TRUE)
auc_fit <- glmQLFTest(fit, coef=c('auc'))
genes <- topTags(auc_fit, n=Inf, p.value = 1)
create_dt(as.data.frame(genes))

```
*Notes: No genes were found to have FDR < 0.05*

#### Top genes

Plotting top genes: 

```{r}
top_genes <- c(genes$table$genes)
top_genes <- top_genes[1:5]

for (gene in top_genes) {
  gene <- as.character(gene)
  y$cpm <- cpm(y, log = T) 
  expression_values <- y$cpm[y$genes$genes==gene,]
  plot<- ggplot(data = NULL, mapping = aes(x=auc, y=expression_values, colour = batch)) + 
    geom_point() +
    theme_linedraw() +
    xlab('AUC values') + 
    ylab('log(cpm)') +
    ggtitle(label=gene)
  print(plot)
}
```

### Cisplatin

#### Design Matrix 

Design matrix used for analysis:

```{r}
counts <- y
auc <- as.numeric(metadata$Cisplatin)
auc_design <- cbind(design,auc)
create_dt(auc_design)
```

#### Fitting GLM

Top genes related to Cisplatin AUC: 

```{r}
counts <- counts %>%
  estimateDisp() %>%
  estimateCommonDisp() %>%
  estimateTagwiseDisp()

fit <- glmQLFit(counts, design = auc_design, robust = TRUE)
auc_fit <- glmQLFTest(fit, coef=c('auc'))
genes <- topTags(auc_fit, n=Inf, p.value = 1)
create_dt(as.data.frame(genes))

```
*Notes: No genes were found to have FDR < 0.05*

#### Top genes

Plotting top genes: 

```{r}
top_genes <- c(genes$table$genes)
top_genes <- top_genes[1:5]

for (gene in top_genes) {
  gene <- as.character(gene)
  y$cpm <- cpm(y, log = T) 
  expression_values <- y$cpm[y$genes$genes==gene,]
  plot<- ggplot(data = NULL, mapping = aes(x=auc, y=expression_values, colour = batch)) + 
    geom_point() +
    theme_linedraw() +
    xlab('AUC values') + 
    ylab('log(cpm)') +
    ggtitle(label=gene)
  print(plot)
}
```

### Docetaxel 

#### Design Matrix 

Design matrix used for analysis:

```{r}
counts <- y
auc <- as.numeric(metadata$Docetaxel)
auc_design <- cbind(design,auc)
create_dt(auc_design)
```

#### Fitting GLM

Top genes related to Docetaxel AUC: 

```{r}
counts <- counts %>%
  estimateDisp() %>%
  estimateCommonDisp() %>%
  estimateTagwiseDisp()

fit <- glmQLFit(counts, design = auc_design, robust = TRUE)
auc_fit <- glmQLFTest(fit, coef=c('auc'))
genes <- topTags(auc_fit, n=Inf, p.value = 1)
create_dt(as.data.frame(genes))

```
*Notes: No genes were found to have FDR < 0.05*

#### Top genes

Plotting top genes: 

```{r}
top_genes <- c(genes$table$genes)
top_genes <- top_genes[1:5]

for (gene in top_genes) {
  gene <- as.character(gene)
  y$cpm <- cpm(y, log = T) 
  expression_values <- y$cpm[y$genes$genes==gene,]
  plot<- ggplot(data = NULL, mapping = aes(x=auc, y=expression_values, colour = batch)) + 
    geom_point() +
    theme_linedraw() +
    xlab('AUC values') + 
    ylab('log(cpm)') +
    ggtitle(label=gene)
  print(plot)
}
```

### Doxorubicin

#### Design Matrix 

Design matrix used for analysis:

```{r}
counts <- y
auc <- as.numeric(metadata$Doxorubicin)
auc_design <- cbind(design,auc)
create_dt(auc_design)
```

#### Fitting GLM

Top genes related to Doxorubicin AUC: 

```{r}
counts <- counts %>%
  estimateDisp() %>%
  estimateCommonDisp() %>%
  estimateTagwiseDisp()

fit <- glmQLFit(counts, design = auc_design, robust = TRUE)
auc_fit <- glmQLFTest(fit, coef=c('auc'))
genes <- topTags(auc_fit, n=Inf, p.value = 1)
create_dt(as.data.frame(genes))

```
*Notes: No genes were found to have FDR < 0.05*

#### Top genes

Plotting top genes: 

```{r}
top_genes <- c(genes$table$genes)
top_genes <- top_genes[1:5]

for (gene in top_genes) {
  gene <- as.character(gene)
  y$cpm <- cpm(y, log = T) 
  expression_values <- y$cpm[y$genes$genes==gene,]
  plot<- ggplot(data = NULL, mapping = aes(x=auc, y=expression_values, colour = batch)) + 
    geom_point() +
    theme_linedraw() +
    xlab('AUC values') + 
    ylab('log(cpm)') +
    ggtitle(label=gene)
  print(plot)
}
```

### Gemcitabine

#### Design Matrix 

Design matrix used for analysis:

```{r}
counts <- y
auc <- as.numeric(metadata$Gemcitabine)
auc_design <- cbind(design,auc)
create_dt(auc_design)
```

#### Fitting GLM

Top genes related to Gemcitabine AUC: 

```{r}
counts <- counts %>%
  estimateDisp() %>%
  estimateCommonDisp() %>%
  estimateTagwiseDisp()

fit <- glmQLFit(counts, design = auc_design, robust = TRUE)
auc_fit <- glmQLFTest(fit, coef=c('auc'))
genes <- topTags(auc_fit, n=Inf, p.value = 1)
create_dt(as.data.frame(genes))

```
*Notes: No genes were found to have FDR < 0.05*

#### Top genes

Plotting top genes: 

```{r}
top_genes <- c(genes$table$genes)
top_genes <- top_genes[1:5]

for (gene in top_genes) {
  gene <- as.character(gene)
  y$cpm <- cpm(y, log = T) 
  expression_values <- y$cpm[y$genes$genes==gene,]
  plot<- ggplot(data = NULL, mapping = aes(x=auc, y=expression_values, colour = batch)) + 
    geom_point() +
    theme_linedraw() +
    xlab('AUC values') + 
    ylab('log(cpm)') +
    ggtitle(label=gene)
  print(plot)
}
```

### Irinotecan

#### Design Matrix 

Design matrix used for analysis:

```{r}
counts <- y
auc <- as.numeric(metadata$Irinotecan)
auc_design <- cbind(design,auc)
create_dt(auc_design)
```

#### Fitting GLM

Top genes related to Irinotecan AUC: 

```{r}
counts <- counts %>%
  estimateDisp() %>%
  estimateCommonDisp() %>%
  estimateTagwiseDisp()

fit <- glmQLFit(counts, design = auc_design, robust = TRUE)
auc_fit <- glmQLFTest(fit, coef=c('auc'))
genes <- topTags(auc_fit, n=Inf, p.value = 1)
create_dt(as.data.frame(genes))

```
*Notes: 1 gene was found to have FDR < 0.05*

#### Top genes

Plotting top genes: 

```{r}
top_genes <- c(genes$table$genes)
top_genes <- top_genes[1:5]

for (gene in top_genes) {
  gene <- as.character(gene)
  y$cpm <- cpm(y, log = T) 
  expression_values <- y$cpm[y$genes$genes==gene,]
  plot<- ggplot(data = NULL, mapping = aes(x=auc, y=expression_values, colour = batch)) + 
    geom_point() +
    theme_linedraw() +
    xlab('AUC values') + 
    ylab('log(cpm)') +
    ggtitle(label=gene)
  print(plot)
}
```

### Mitomycin

#### Design Matrix 

Design matrix used for analysis:

```{r}
counts <- y
auc <- as.numeric(metadata$`Mitomycin C`)
auc_design <- cbind(design,auc)
create_dt(auc_design)
```

#### Fitting GLM

Top genes related to Mitomycin AUC: 

```{r}
counts <- counts %>%
  estimateDisp() %>%
  estimateCommonDisp() %>%
  estimateTagwiseDisp()

fit <- glmQLFit(counts, design = auc_design, robust = TRUE)
auc_fit <- glmQLFTest(fit, coef=c('auc'))
genes <- topTags(auc_fit, n=Inf, p.value = 1)
create_dt(as.data.frame(genes))

```
*Notes: No genes were found to have FDR < 0.05*

#### Top genes

Plotting top genes: 

```{r}
top_genes <- c(genes$table$genes)
top_genes <- top_genes[1:5]

for (gene in top_genes) {
  gene <- as.character(gene)
  y$cpm <- cpm(y, log = T) 
  expression_values <- y$cpm[y$genes$genes==gene,]
  plot<- ggplot(data = NULL, mapping = aes(x=auc, y=expression_values, colour = batch)) + 
    geom_point() +
    theme_linedraw() +
    xlab('AUC values') + 
    ylab('log(cpm)') +
    ggtitle(label=gene)
  print(plot)
}
```

### Oxaliplatin

#### Design Matrix 

Design matrix used for analysis:

```{r}
counts <- y
auc <- as.numeric(metadata$Oxaliplatin)
auc_design <- cbind(design,auc)
create_dt(auc_design)
```

#### Fitting GLM

Top genes related to Oxaliplatin AUC: 

```{r}
counts <- counts %>%
  estimateDisp() %>%
  estimateCommonDisp() %>%
  estimateTagwiseDisp()

fit <- glmQLFit(counts, design = auc_design, robust = TRUE)
auc_fit <- glmQLFTest(fit, coef=c('auc'))
genes <- topTags(auc_fit, n=Inf, p.value = 1)
create_dt(as.data.frame(genes))

```
*Notes: No genes were found to have FDR < 0.05*

#### Top genes

Plotting top genes: 

```{r}
top_genes <- c(genes$table$genes)
top_genes <- top_genes[1:5]

for (gene in top_genes) {
  gene <- as.character(gene)
  y$cpm <- cpm(y, log = T) 
  expression_values <- y$cpm[y$genes$genes==gene,]
  plot<- ggplot(data = NULL, mapping = aes(x=auc, y=expression_values, colour = batch)) + 
    geom_point() +
    theme_linedraw() +
    xlab('AUC values') + 
    ylab('log(cpm)') +
    ggtitle(label=gene)
  print(plot)
}
```

### Paclitaxel

#### Design Matrix 

Design matrix used for analysis:

```{r}
counts <- y
auc <- as.numeric(metadata$Paclitaxel)
auc_design <- cbind(design,auc)
create_dt(auc_design)
```

#### Fitting GLM

Top genes related to Paclitaxel AUC: 

```{r}
counts <- counts %>%
  estimateDisp() %>%
  estimateCommonDisp() %>%
  estimateTagwiseDisp()

fit <- glmQLFit(counts, design = auc_design, robust = TRUE)
auc_fit <- glmQLFTest(fit, coef=c('auc'))
genes <- topTags(auc_fit, n=Inf, p.value = 1)
create_dt(as.data.frame(genes))

```
*Notes: No genes were found to have FDR < 0.05*

#### Top genes

Plotting top genes: 

```{r}
top_genes <- c(genes$table$genes)
top_genes <- top_genes[1:5]

for (gene in top_genes) {
  gene <- as.character(gene)
  y$cpm <- cpm(y, log = T) 
  expression_values <- y$cpm[y$genes$genes==gene,]
  plot<- ggplot(data = NULL, mapping = aes(x=auc, y=expression_values, colour = batch)) + 
    geom_point() +
    theme_linedraw() +
    xlab('AUC values') + 
    ylab('log(cpm)') +
    ggtitle(label=gene)
  print(plot)
}
```

### Topotecan

#### Design Matrix

Design matrix used for analysis:

```{r}
counts <- y
auc <- as.numeric(metadata$Topotecan)
auc_design <- cbind(design,auc)
create_dt(auc_design)
```

#### Fitting GLM

Top genes related to Topotecan AUC: 

```{r}
counts <- counts %>%
  estimateDisp() %>%
  estimateCommonDisp() %>%
  estimateTagwiseDisp()

fit <- glmQLFit(counts, design = auc_design, robust = TRUE)
auc_fit <- glmQLFTest(fit, coef=c('auc'))
genes <- topTags(auc_fit, n=Inf, p.value = 1)
create_dt(as.data.frame(genes))

```
*Notes: No genes were found to have FDR < 0.05*

#### Top genes

Plotting top genes: 

```{r}
top_genes <- c(genes$table$genes)
top_genes <- top_genes[1:5]

for (gene in top_genes) {
  gene <- as.character(gene)
  y$cpm <- cpm(y, log = T) 
  expression_values <- y$cpm[y$genes$genes==gene,]
  plot<- ggplot(data = NULL, mapping = aes(x=auc, y=expression_values, colour = batch)) + 
    geom_point() +
    theme_linedraw() +
    xlab('AUC values') + 
    ylab('log(cpm)') +
    ggtitle(label=gene)
  print(plot)
}
```


## {-}
